---
title: "ACS data on counties surrounding Atlanta"
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r, include = FALSE}
setwd("/mnt/c/Users/adavi/Documents/comp-journal/ajc-satellite/data-findings/")
library(tidyverse)
library(tidycensus)
library(lubridate)
library(zoo)
library(sf)
```

# Geographic units and time period(s) considered

Throughout, we will consider different geographic units, corresponding to
various collections of counties, and other unified geographical areas corresponding
to different parts of the landscape.

```{r}
core_counties <- c(
    "Fulton", "DeKalb", "Gwinnett", "Cobb", "Clayton"
)

arc_counties <- c(
    "Fulton", "DeKalb", "Gwinnett", "Cobb", "Clayton", "Forsyth",
    "Cherokee", "Douglas", "Fayette", "Henry", "Rockdale"
)

atlanta_msa_1 <- c(
    'Fulton', 'Gwinnett',
    'Cobb', 'DeKalb', 'Clayton',
    'Cherokee', 'Forsyth',
    'Henry', 'Paulding', 'Coweta',
    'Douglas', 'Fayette', 'Carroll',
    'Newton', 'Bartow', 'Walton', 
    'Rockdale'
)

atlanta_msa_2 <- c(
    'Barrow', 'Spalding', 'Pickens', 
    'Haralson', 'Dawson', 'Butts', 
    'Meriwether', 'Morgan',
    'Pike', 'Lamar', 'Jasper', 'Heard'
)

atlanta_msa <- c(atlanta_msa_1, atlanta_msa_2)
```

Throughout, we will frequently use ACS data; for ACS 1 year estimates, we use data
from 2009 to 2019 (we note that [2020 data has not been released](
    https://www.census.gov/newsroom/press-releases/2021/changes-2020-acs-1-year.html
) as a result of the COVID pandemic adversely affecting data collection), and for
ACS 5 year estimates, we use data from 2009 to 2020.

```{r}
years_acs1 <- lst(
    2009, 2010, 2011, 2012, 2013, 2014,
    2015, 2016, 2017, 2018, 2019
)

years_acs5 <- lst(
    2009, 2010, 2011, 2012, 2013, 2014, 
    2015, 2016, 2017, 2018, 2019, 2020
)
```

## Defining inside and outside the perimeter

Here, we will consider the I-285 ringroad as defining the "perimeter"
of the "main" part of Atlanta, and will consider anything inside of this
as being "inside the perimeter" (or ITP), and anything outside as
"outside the perimeter" (or OTP). The shape file for the ringroad is available
[online](https://opendata.atlantaregional.com/datasets/929e158242f24c9fb260606526b7d80e).

```{r}
highways <- sf::st_read("../data/georgia-highways.geojson")

it_285 <- highways %>%
    filter(str_detect(ROAD_NAME, "285")) %>%
    select(ROAD_NAME, geometry)

it_285 <- it_285[c(1, 3), ]
itp <- st_union(st_convex_hull(it_285))
```

As is visible from the plot below, the ringroad does not necessarily
line up nicely with the county divisions within Georgia: the perimeter
is highlighted in red, with the regular counties highlighted in black.
However, the census tracts (in grey) do line up more nicely with the
perimeter, albeit not exactly. 

```{r}
arc_counties_shp <- sf::st_read("../ga-counties/Counties_Georgia.shp") %>%
    filter(NAME10 %in% arc_counties) %>%
    select(NAME10, COUNTYFP10, geometry)

arc_tracts_shp <- sf::st_read("../ga-tracts-2020/tl_2020_13_tract.shp") %>%
    filter(COUNTYFP %in% arc_counties_shp$COUNTYFP10)

ggplot() +
    geom_sf(data = arc_tracts_shp, colour = "grey", fill = NA) +
    geom_sf(data = arc_counties_shp, colour = "black", fill = NA) +
    geom_sf(data = itp, colour = "red", fill = NA)
```

Ultimately, we want to try and assign each census tract to being ITP
or OTP. Firstly, to handle this, we count the number of tracts (where we
use the 2020 Census Tracts) which intersect with the "perimeter" itself.

```{r}
ics <- st_intersects(
    st_transform(st_boundary(itp), crs = 4269),
    arc_tracts_shp
)

print(
    "Number of census tracts intersecting the interstate: ",
    length(ics[[1]])
)
```

As we can see, a small but non-zero number fall inside of this category, and
so we will keep track of:

- census tracts inside the perimeter (ITP),
- census tracts on the perimeter (TP), and
- census tracts outside the perimeter (OTP).

```{r}
arc_tracts_shp <- arc_tracts_shp %>% 
    select(NAME, COUNTYFP, TRACTCE, geometry) %>%
    mutate(
        tp = st_intersects(
            st_transform(st_boundary(itp), crs = 4269),
            geometry, sparse = FALSE
        )[1, ],
        itp = st_within(
            geometry,
            st_transform(itp, crs = 4269), sparse = FALSE
        )[, 1],
        otp = !tp & !itp
    )

arc_tracts_shp %>% 
    st_drop_geometry() %>%
    summarize(
        tp = sum(tp), itp = sum(itp), otp = sum(otp)
    )
```

```{r}
arc_tracts_shp2 <- arc_tracts_shp %>%
    pivot_longer(
        cols = c(tp, itp, otp),
        names_to = "class",
        values_to = "in_class"
    ) %>%
    filter(in_class == TRUE) %>%
    select(-in_class)

ggplot() +
    geom_sf(data = arc_tracts_shp2, colour = "grey", mapping = aes(fill = class)) +
    geom_sf(data = arc_counties_shp, colour = "black", fill = NA) +
    geom_sf(data = itp, colour = "red", fill = NA)
```

# Household income distribution in the Atlanta metro area between 2009 and 2019

```{r, include = FALSE}
v11 <- load_variables(2011, "acs1", cache = TRUE)

income_variables <- paste0("B19001_0", str_pad(1:17, 2, pad = "0"))

income_var_labels <- v11 %>%
    filter(name %in% income_variables) %>%
    select(label) %>% as_vector()

income_var_labels <- sapply(
    str_split(income_var_labels, "!!"), tail, 1
)

income_var_list <- setNames(as.list(income_variables), income_var_labels)

df_income_distribution <- map_dfr(
    years_acs1,
    ~ get_acs(
        geography = "county",
        state = "GA",
        variables = income_var_list,
        year = .x,
        county = arc_counties,
        survey = "acs1",
        geometry = FALSE
    ),
    .id = "year") %>%
    mutate(
        county = gsub("^(.*?),.*", "\\1", NAME)
    ) %>% select(-NAME) %>% relocate(county, .after = GEOID)
```

ACS 1-year data from 2009 to 2019, showing the number of households (with
error estimates) which lie in particular income bands across the ARC counties.

```{r}
df_income_distribution
```

## How has the income distribution changed over the years across the Atlanta area?

To answer this, we summarize over all the counties the number of households in
each income bracket, along with the total number, and do so for each year. The margin
of errors give 90% confidence intervals of each estimate; to get the margin of error for
a sum of variables, we sum the square of the margin of error across all the variables,
and then take the square root. For this analysis, we look at all the counties
which belong to the Altanta Regional Commission. 

### Looking at number of households

```{r}
df_income_dist_summary <- df_income_distribution %>%
    select(year, county, variable, estimate, moe) %>%
    group_by(year, variable) %>%
    summarize(
        estimate = sum(estimate),
        moe = sqrt(sum(moe**2))
    ) %>%
    ungroup() %>%
    mutate(
        variable = factor(variable, ordered = TRUE, levels = income_var_labels)
    ) %>%
    filter(variable != "Total")

df_income_dist_summary

ggplot(df_income_dist_summary, aes(x = variable, y = estimate)) + 
    facet_wrap(~ year) +
    geom_col() +
    geom_errorbar(aes(ymin = estimate - moe, ymax = estimate + moe)) +
    labs(
        x = "Income band", y = "Number of households",
        title = "Distribution of Household Income in the Atlanta area Between 2009 and 2019"
    ) + coord_flip()
```

### Looking at the proportion of households

```{r}
df_income_prop_summary <- df_income_distribution %>%
    select(year, county, variable, estimate, moe) %>%
    group_by(year, variable) %>%
    summarize(
        estimate = sum(estimate),
        moe = sqrt(sum(moe**2))
    ) %>%
    ungroup() %>%
    mutate(
        variable = factor(variable, ordered = TRUE, levels = income_var_labels)
    ) %>%
    filter(variable != "Total") %>%
    group_by(year) %>%
    mutate(
        moe = moe / sum(estimate),
        estimate = estimate / sum(estimate),
    )

df_income_prop_summary

ggplot(df_income_prop_summary, aes(x = variable, y = estimate)) + 
    facet_wrap(~ year) +
    geom_col() +
    geom_errorbar(aes(ymin = estimate - moe, ymax = estimate + moe)) +
    labs(
        x = "Income band", y = "Proportion of households",
        title = "Distribution of Household Income in the Atlanta area Between 2009 and 2019"
    ) + coord_flip()
```

### Change between 2009 and 2019

Looking at the number of households in each income band at 2009 and 2019, we can see
that there has been a large increase in the number of households earning $100k or more,
while the number of households earning less than $50k has mostly stayed constant. 

```{r}
df_income_dist_summary %>%
    filter(year %in% c(2009, 2019)) %>%
    ggplot(aes(x = variable, y = estimate, fill = year)) +
    geom_col(position = "dodge") +
    geom_errorbar(
        aes(ymin = estimate - moe, ymax = estimate + moe), 
        position = "dodge"
    ) +
    labs(
        y = "Proportion of households", x = "Income band",
        title = "Distribution of Household Income in the Atlanta area (2009 and 2019)"
    ) + 
    coord_flip() 
```

### Changes across counties over time

```{r}
df_income_distribution %>%
    select(year, county, variable, estimate, moe) %>%
    filter(variable != "Total") %>%
    group_by(year, county) %>%
    mutate(estimate = estimate / sum(estimate)) %>%
    ungroup() %>%
    mutate(
        variable = factor(variable, ordered = TRUE, levels = income_var_labels)
    ) %>%
    ggplot(aes(x = year, y = estimate, fill = forcats::fct_rev(variable))) +
    facet_wrap(~ county) +
    geom_col() + coord_flip() +
    labs(
        x = "Year", y = "Proportion of Households", fill = "Income Band",
        title = "Distribution of Household Income across Counties and Year"
    ) +
    theme(legend.position = "bottom")
```

## How has the income distribution changed ITP and OTP from 2009 to 2019?

```{r}
df_income_tracts <- map_dfr(
    lst(2009, 2020),
    ~ get_acs(
        geography = "tract",
        state = "GA",
        variables = income_var_list,
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = TRUE
    ),
    .id = "year")
```