---
title: "ACS and satellite data on Atlanta and the wider metropoliton area"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    theme: flatly
    toc_float:
        collapsed: false
    highlight: pygments
    fig_caption: true
    fig_width: 10
    fig_height: 8
    df_print: paged
    self_contained: false
    code_folding: hide
---

```{r, include = FALSE}
setwd("/mnt/c/Users/adavi/Documents/comp-journal/ajc-satellite/data-findings/")
options(width = 1200)
options(scipen = 100000000)
options(tigris_use_cache = TRUE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(tidyverse)
library(tidycensus)
library(lubridate)
library(zoo)
library(sf)
library(viridis)
```

# Geographic units and time period(s) considered

Throughout, we will consider different geographic units, corresponding to
various collections of counties, and other unified geographical areas corresponding
to different parts of the landscape.

```{r}
core_counties <- c(
    "Fulton", "DeKalb", "Gwinnett", "Cobb", "Clayton"
)

arc_counties <- c(
    "Fulton", "DeKalb", "Gwinnett", "Cobb", "Clayton", "Forsyth",
    "Cherokee", "Douglas", "Fayette", "Henry", "Rockdale"
)

atlanta_msa_1 <- c(
    'Fulton', 'Gwinnett',
    'Cobb', 'DeKalb', 'Clayton',
    'Cherokee', 'Forsyth',
    'Henry', 'Paulding', 'Coweta',
    'Douglas', 'Fayette', 'Carroll',
    'Newton', 'Bartow', 'Walton', 
    'Rockdale'
)

atlanta_msa_2 <- c(
    'Barrow', 'Spalding', 'Pickens', 
    'Haralson', 'Dawson', 'Butts', 
    'Meriwether', 'Morgan',
    'Pike', 'Lamar', 'Jasper', 'Heard'
)

atlanta_msa <- c(atlanta_msa_1, atlanta_msa_2)
```

Throughout, we will frequently use ACS data; for ACS 1 year estimates, we use data
from 2009 to 2019 (we note that [2020 data has not been released](
    https://www.census.gov/newsroom/press-releases/2021/changes-2020-acs-1-year.html
) as a result of the COVID pandemic adversely affecting data collection), and for
ACS 5 year estimates, we use data in the range
from the 2006-to-2010 estimates to the 2016-to-2020.

```{r}
years_acs1 <- lst(
    2009, 2010, 2011, 2012, 2013, 2014,
    2015, 2016, 2017, 2018, 2019
)

years_acs5 <- lst(
    2010, 2011, 2012, 2013, 2014, 
    2015, 2016, 2017, 2018, 2019, 2020
)
```

## Defining inside and outside the perimeter

Here, we will consider the I-285 ringroad as defining the "perimeter"
of the "main" part of Atlanta, and will consider anything inside of this
as being "inside the perimeter" (or ITP), and anything outside as
"outside the perimeter" (or OTP). The shape file for the ringroad is available
[online](https://opendata.atlantaregional.com/datasets/929e158242f24c9fb260606526b7d80e).

```{r}
highways <- sf::st_read("../data/georgia-highways.geojson")

it_285 <- highways %>%
    filter(str_detect(ROAD_NAME, "285")) %>%
    select(ROAD_NAME, geometry)

it_285 <- it_285[c(1, 3), ]
itp <- st_union(st_convex_hull(it_285))
```

As is visible from the plot below, the ringroad does not necessarily
line up nicely with the county divisions within Georgia: the perimeter
is highlighted in red, with the regular counties highlighted in black.
However, the census tracts (in grey) do line up more nicely with the
perimeter, albeit not exactly. 

```{r}
arc_counties_shp <- sf::st_read("../ga-counties/Counties_Georgia.shp") %>%
    filter(NAME10 %in% arc_counties) %>%
    select(NAME10, COUNTYFP10, geometry)

arc_tracts_shp <- sf::st_read("../ga-tracts-2019/tl_2019_13_tract.shp") %>%
    filter(COUNTYFP %in% arc_counties_shp$COUNTYFP10)

highway_arc <- st_intersection(highways, arc_counties_shp)

ggplot() +
    geom_sf(data = arc_tracts_shp, colour = "grey", fill = NA) +
    geom_sf(data = arc_counties_shp, colour = "black", fill = NA) +
    geom_sf(data = itp, colour = "red", fill = NA)
```

Ultimately, we want to try and assign each census tract to being ITP
or OTP. Firstly, to handle this, we count the number of tracts (where we
use the 2020 Census Tracts) which intersect with the "perimeter" itself.

```{r}
ics <- st_intersects(
    st_transform(st_boundary(itp), crs = 4269),
    arc_tracts_shp
)

print(paste(
    "Number of census tracts intersecting the interstate:",
    length(ics[[1]])
))
```

As we can see, a small but non-zero number fall inside of this category;
we will count these tracts as being inside of the perimeter.

```{r}
arc_tracts_shp <- arc_tracts_shp %>% 
    select(NAME, COUNTYFP, TRACTCE, geometry) %>%
    mutate(
        tp = st_intersects(
            st_transform(st_boundary(itp), crs = 4269),
            geometry, sparse = FALSE
        )[1, ],
        itp = st_within(
            geometry,
            st_transform(itp, crs = 4269), sparse = FALSE
        )[, 1],
        itp = tp | itp,
        otp = !itp
    ) %>% select(-tp)

arc_tracts_shp %>% 
    st_drop_geometry() %>%
    summarize(
        itp = sum(itp), otp = sum(otp)
    )
```

```{r}
arc_tracts_shp2 <- arc_tracts_shp %>%
    pivot_longer(
        cols = c(itp, otp),
        names_to = "class",
        values_to = "in_class"
    ) %>%
    filter(in_class == TRUE) %>%
    select(-in_class)

ggplot() +
    geom_sf(
        data = arc_tracts_shp2, colour = "grey",
        mapping = aes(fill = class)
    ) +
    geom_sf(data = arc_counties_shp, colour = "black", fill = NA) +
    geom_sf(data = itp, colour = "red", fill = NA) +
    coord_sf(xlim = c(-84.6, -84), ylim = c(33.4, 34), expand = TRUE)
```

# Household income distribution in the Atlanta metro area between 2009 and 2019

```{r, include = FALSE}
v11 <- load_variables(2011, "acs1", cache = TRUE)

income_variables <- paste0("B19001_0", str_pad(1:17, 2, pad = "0"))

income_var_labels <- v11 %>%
    filter(name %in% income_variables) %>%
    select(label) %>% as_vector()

income_var_labels <- sapply(
    str_split(income_var_labels, "!!"), tail, 1
)

income_var_list <- setNames(as.list(income_variables), income_var_labels)

df_income_distribution <- map_dfr(
    years_acs1,
    ~ get_acs(
        geography = "county",
        state = "GA",
        variables = income_var_list,
        year = .x,
        county = arc_counties,
        survey = "acs1",
        geometry = FALSE
    ),
    .id = "year") %>%
    mutate(
        county = gsub("^(.*?),.*", "\\1", NAME)
    ) %>% select(-NAME) %>% relocate(county, .after = GEOID)

df_income_distribution <- df_income_distribution %>%
    mutate(variable = fct_collapse(variable, 
            "Less than $50,000" = c(
                "Less than $10,000", "$10,000 to $14,999", 
                "$15,000 to $19,999", "$20,000 to $24,999",
                "$25,000 to $29,999", "$30,000 to $34,999",
                "$35,000 to $39,999", "$40,000 to $44,999",
                "$45,000 to $49,999"
            ),
            "$50,000 to $74,999" = c("$50,000 to $59,999", "$60,000 to $74,999"),
            "$75,000 to $99,999" = c("$75,000 to $99,999"),
            "$100,000 to $149,999" = c(
                "$100,000 to $124,999", "$125,000 to $149,999"
            ),
            "$150,000 to $199,999" = c("$150,000 to $199,999"),
            "$200,000 or more" = c("$200,000 or more")
        )) %>%
    group_by(year, GEOID, county, variable) %>%
    summarize(
        estimate = sum(estimate),
        moe = sqrt(sum(moe**2))
    )

income_var_labels <- c(
    "Less than $50,000", "$50,000 to $74,999",
    "$75,000 to $99,999", "$100,000 to $149,999",
    "$150,000 to $199,999", "$200,000 or more"
)
```

ACS 1-year data from 2009 to 2019, showing the number of households (with
error estimates) which lie in particular income bands across the ARC counties.

```{r}
df_income_distribution
write_csv(
    df_income_distribution,
    "data-files/income_distribution_by_year_county.csv"
)
```

## How has the income distribution changed over the years across the Atlanta area?

To answer this, we summarize over all the counties the number of households in
each income bracket, along with the total number, and do so for each year. The margin
of errors give 90% confidence intervals of each estimate; to get the margin of error for
a sum of variables, we sum the square of the margin of error across all the variables,
and then take the square root. For this analysis, we look at all the counties
which belong to the Altanta Regional Commission. 

### Looking at number of households

From the plots below, we can see that the majority of the shift in the
income distribution has been a large increase in the number of households
earning between $150,000 and $199,999 and more than $200,000 (by almost 100,000 from 2009 to 2019 in both groups), with a
similar drop of almost 100,000 households earning less than $50,000.

```{r}
df_income_dist_summary <- df_income_distribution %>%
    select(year, county, variable, estimate, moe) %>%
    group_by(year, variable) %>%
    summarize(
        estimate = sum(estimate),
        moe = sqrt(sum(moe**2))
    ) %>%
    ungroup() %>%
    mutate(
        variable = factor(variable, ordered = TRUE,
        levels = income_var_labels)
    ) %>%
    filter(variable != "Total")

df_income_dist_summary
write_csv(
    df_income_dist_summary,
    "data-files/income_distribution_by_year.csv"
)

require(scales)

ggplot(df_income_dist_summary, aes(x = variable, y = estimate)) + 
    facet_wrap(~ year) +
    geom_col() +
    geom_errorbar(aes(ymin = estimate - moe, ymax = estimate + moe)) +
    labs(
        x = "Income band", y = "Number of households",
        title = "Distribution of Household Income in the Atlanta area Between 2009 and 2019"
    ) + scale_y_continuous(labels = comma) +
    coord_flip()
```

```{r}
df_income_dist_summary %>%
    mutate(year = as.Date(paste(year, 1, 1, sep = "-"))) %>%
    ggplot(aes(
        x = year, y = estimate, col = variable, fill = variable,
        ymin = estimate - moe, ymax = estimate + moe)
    ) + geom_point(show.legend = FALSE) + geom_line() + 
    geom_ribbon(alpha = 0.2, show.legend = FALSE) +
    labs(
        x = "Year", y = "Number of households",
        col = "Income band"
    ) +
    theme(legend.position = "bottom")
```

### Looking at the proportion of households

```{r}
df_income_prop_summary <- df_income_distribution %>%
    select(year, county, variable, estimate, moe) %>%
    group_by(year, variable) %>%
    summarize(
        estimate = sum(estimate),
        moe = sqrt(sum(moe**2))
    ) %>%
    ungroup() %>%
    mutate(
        variable = factor(variable, ordered = TRUE, levels = income_var_labels)
    ) %>%
    filter(variable != "Total") %>%
    group_by(year) %>%
    mutate(
        moe = moe / sum(estimate),
        estimate = estimate / sum(estimate),
    )

df_income_prop_summary

ggplot(df_income_prop_summary, aes(x = variable, y = estimate)) + 
    facet_wrap(~ year) +
    geom_col() +
    geom_errorbar(aes(ymin = estimate - moe, ymax = estimate + moe)) +
    labs(
        x = "Income band", y = "Proportion of households",
        title = "Distribution of Household Income in the Atlanta area Between 2009 and 2019"
    ) + scale_y_continuous(labels = comma) + coord_flip()
```

### Overall change between 2009 and 2019

Here we focus particularly on the change between the years in 2009 and
2019.

```{r}
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

df_income_dist_summary %>%
    filter(year %in% c(2009, 2019)) %>%
    mutate(year = factor(year)) %>%
    ggplot(aes(x = variable, y = estimate, fill = year)) +
    geom_col(position = "dodge") +
    geom_errorbar(
        aes(ymin = estimate - moe, ymax = estimate + moe), 
        position = "dodge"
    ) + scale_fill_manual(values = cbPalette) +
    geom_text(
        aes(label = year, y = 10000), 
        position = position_dodge(width = 1),
        size = 5, hjust = 'left',
        col = "white"
    ) + 
    labs(
        y = "Number of households", x = "Income band",
        title = "Distribution of Household Income in the Atlanta area (2009 and 2019)", fill = "Year",
        subtitle = "Data from the 1-year ACS from 2009 and 2019"
    ) + scale_y_continuous(labels = comma) +
    coord_flip() + theme(legend.position = "none")

ggsave("img-files/income_distribution_by_year.pdf")
```

### Looking at the changes in each county

```{r}
df_income_distribution %>%
    select(year, county, variable, estimate, moe) %>%
    filter(variable != "Total") %>%
    mutate(
        variable = factor(variable, ordered = TRUE, levels = income_var_labels)
    ) %>%
    ggplot(aes(x = year, y = estimate, fill = forcats::fct_rev(variable))) +
    facet_wrap(~ county, scales = "free") +
    geom_col() + coord_flip() +
    labs(
        x = "Year", y = "Number of Households", fill = "Income Band",
        title = "Distribution of Household Income across Counties and Year",
        subtitle = "Data from the 1-year ACS from 2009 to 2019"
    ) +
    theme(legend.position = "bottom")
```

## How has the income distribution changed ITP and OTP from 2009 to 2019?

From the plot below, we can see that the number of households inside
of the perimeter earning less than $50,000 has decreased by approximately
25000, whereas that outside the perimeter has stayed roughly constant. 

```{r, results = 'hide'}
df_income_tracts <- map_dfr(
    lst(2009, 2019),
    ~ get_acs(
        geography = "tract",
        state = "GA",
        variables = income_var_list,
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = FALSE
    ),
    .id = "year") %>%
    mutate(
        tract = gsub("^(.*?),.*", "\\1", NAME),
        county = gsub("^(.*?), (.*?),.*", "\\2", NAME)
    ) %>% 
    select(-NAME) %>%
    relocate(tract, county, .after = GEOID)

df_tracts <- map_dfr(
    lst(2009, 2019),
    ~ get_acs(
        geography = "tract",
        state = "GA",
        variables = "B19001_001",
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = TRUE
    ),
    .id = "year") %>% 
    select(year, GEOID, geometry) %>%
    mutate(
        tp = st_intersects(
            st_transform(st_boundary(itp), crs = 4269),
            geometry, sparse = FALSE
        )[1, ],
        itp = st_within(
            geometry,
            st_transform(itp, crs = 4269), sparse = FALSE
        )[, 1],
        itp = tp | itp,
        otp = !itp
    ) %>% select(-tp) %>%
    pivot_longer(
        cols = c(itp, otp),
        names_to = "class",
        values_to = "in_class"
    ) %>%
    filter(in_class == TRUE) %>%
    select(-in_class)

df_tracts_nogeom <- df_tracts %>% st_drop_geometry()

df_income_by_perimeter <- df_income_tracts %>%
    filter(variable != "Total") %>%
    left_join(df_tracts_nogeom) %>%
    select(year, class, variable, estimate, moe) %>%
    mutate(variable = fct_collapse(variable, 
            "Less than $50,000" = c(
                "Less than $10,000", "$10,000 to $14,999", 
                "$15,000 to $19,999", "$20,000 to $24,999",
                "$25,000 to $29,999", "$30,000 to $34,999",
                "$35,000 to $39,999", "$40,000 to $44,999",
                "$45,000 to $49,999"
            ),
            "$50,000 to $74,999" = c("$50,000 to $59,999", "$60,000 to $74,999"),
            "$75,000 to $99,999" = c("$75,000 to $99,999"),
            "$100,000 to $149,999" = c(
                "$100,000 to $124,999", "$125,000 to $149,999"
            ),
            "$150,000 to $199,999" = c("$150,000 to $199,999"),
            "$200,000 or more" = c("$200,000 or more")
        )) %>%
    group_by(year, class, variable) %>%
    summarize(
        estimate = sum(estimate),
        moe = sqrt(sum(moe**2))
    ) %>%
    mutate(
        variable = factor(variable, 
            ordered = TRUE, levels = income_var_labels
        ),
        class = recode_factor(factor(class), 
            itp = "Inside the perimeter", otp = "Outside the perimeter"
        )
    )

df_income_by_perimeter
write_csv(
    df_income_by_perimeter,
    "data-files/income_distribution_inside_outside_perimeter.csv"
)

df_income_by_perimeter %>%
    ggplot(aes(x = variable, y = estimate, fill = year)) +
    geom_col(position = "dodge") + 
    geom_errorbar(
        aes(ymin = estimate - moe, ymax = estimate + moe),
        position = "dodge"
    ) + scale_fill_manual(values = cbPalette) +
    geom_text(
        aes(label = year, y = 10000), 
        position = position_dodge(width = 1),
        size = 5, hjust = 'left',
        col = "#000000"
    ) +
    facet_grid(~ class, scales = "free") +
    labs(
        x = "Income band", y = "Number of households", fill = "Year",
        title = "Distribution of household income in the Atlanta inside and outside the perimeter",
        subtitle = "Data from the 1-year ACS from 2009 and 2019"
    ) +
    coord_flip() + 
    theme(legend.position = "none")
```

## How has the income distribution changed across census tracts over time?

To compare the income distribution across tracts over time, for each census tract,
we will look at the proportion of the population in the different income bands, and
then compare these between 2010 and 2019. (These time periods were chosen so that
there are no issues in switching between different census tracts.) To do so, we 
will compute the total variation distance between the two distributions;
formally, if we have $K$ different possible values, and we have two sets of proportions
labelled $p(i)$ and $q(i)$ (where $i$ can take on $1$, $2$, ..., or $K$), then we compute
the value
$$ \frac{1}{2} \times \Big( |p(1) - q(1)| + |p(2) - q(2)| + \cdots + |p(K) - q(K)| \Big), $$
and use this as a measure of how different $p$ and $q$ are. If this distance is zero,
then the two distributions are identical; the closer the distance is to one, the greater
the difference between the distributions. To try and observe what
is happening in relation to some other geographical structures, the major highways
beyond that of the interstate have also been added to the plots below
as dashed black lines (with the county boundaries given as solid black
lines, and the census tract boundaries as solid grey lines).

```{r, results = 'hide'}
df_income_tracts <- map_dfr(
    lst(2010, 2019),
    ~ get_acs(
        geography = "tract",
        state = "GA",
        variables = income_var_list,
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = FALSE
    ),
    .id = "year") %>%
    mutate(
        tract = gsub("^(.*?),.*", "\\1", NAME),
        county = gsub("^(.*?), (.*?),.*", "\\2", NAME)
    ) %>% 
    select(-NAME) %>%
    relocate(tract, county, .after = GEOID)

df_tracts <- map_dfr(
    lst(2010),
    ~ get_acs(
        geography = "tract",
        state = "GA",
        variables = "B19001_001",
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = TRUE
    ),
    .id = "year") %>% 
    select(GEOID, geometry) %>%
    mutate(
        tp = st_intersects(
            st_transform(st_boundary(itp), crs = 4269),
            geometry, sparse = FALSE
        )[1, ],
        itp = st_within(
            geometry,
            st_transform(itp, crs = 4269), sparse = FALSE
        )[, 1],
        itp = tp | itp,
        otp = !itp
    ) %>% select(-tp) %>%
    pivot_longer(
        cols = c(itp, otp),
        names_to = "class",
        values_to = "in_class"
    ) %>%
    filter(in_class == TRUE) %>%
    select(-in_class)

df_total_households_tracts <- df_income_tracts %>%
    filter(variable == 'Total') %>% 
    select(year, GEOID, estimate, moe)

df_total_households_tracts2 <- df_total_households_tracts %>% 
    filter(year == '2019') %>% select(GEOID, estimate)

df_income_change_long <- df_income_tracts %>%
    filter(variable != "Total") %>%
    select(year, GEOID, variable, estimate) %>%
    mutate(variable = fct_collapse(variable, 
            "Less than $50,000" = c(
                "Less than $10,000", "$10,000 to $14,999", 
                "$15,000 to $19,999", "$20,000 to $24,999",
                "$25,000 to $29,999", "$30,000 to $34,999",
                "$35,000 to $39,999", "$40,000 to $44,999",
                "$45,000 to $49,999"
            ),
            "$50,000 to $74,999" = c("$50,000 to $59,999", "$60,000 to $74,999"),
            "$75,000 to $99,999" = c("$75,000 to $99,999"),
            "$100,000 to $149,999" = c(
                "$100,000 to $124,999", "$125,000 to $149,999"
            ),
            "$150,000 to $199,999" = c("$150,000 to $199,999"),
            "$200,000 or more" = c("$200,000 or more")
        )) %>%
    group_by(year, GEOID, variable) %>%
    summarize(estimate = sum(estimate)) %>%
    ungroup() %>%
    mutate(
        variable = factor(variable, ordered = TRUE, levels = income_var_labels)
    )

df_income_change_tracts <- df_income_change_long %>%
    group_by(year, GEOID) %>%
    mutate(estimate = estimate / sum(estimate)) %>%
    pivot_wider(
        names_from = year,
        names_prefix = "estimate_",
        values_from = estimate
    ) %>% 
    mutate(
        diff = abs(estimate_2019 - estimate_2010)
    ) %>%
    group_by(GEOID) %>%
    summarize(income_dist_change = 0.5*sum(diff)) %>%
    left_join(df_tracts)

df_income_change_overall <- df_income_change_long %>%
    group_by(year) %>%
    mutate(estimate = estimate / sum(estimate)) %>%
    pivot_wider(
        names_from = year,
        names_prefix = "estimate_",
        values_from = estimate
    ) %>% 
    mutate(
        diff = abs(estimate_2010 - estimate_2019)
    ) %>%
    summarize(income_dist_change = 0.5 * sum(diff))

plot(
    df_total_households_tracts2$estimate, df_income_change_tracts$income_dist_change,
    col = factor(df_tracts$class),
    xlab = "Number of households in census tract in 2019",
    ylab = "Total variation of income distribution change in census tract",
    main = "Number of households vs income change across census tracts",
    sub = "(Horizontal red line corresponds to level of income change across all Atlanta)"
) 
abline(h = as.numeric(df_income_change_overall), col = "red", lwd = 3)

df_income_change_tracts %>% 
    mutate(
        class = recode_factor(factor(class), 
            itp = "Inside the perimeter", otp = "Outside the perimeter"
    )) %>%
    select(-geometry) %>%
    ggplot(aes(x = income_dist_change, after_stat(density))) +
    geom_histogram() + facet_wrap( ~ class) +
    geom_vline(
        xintercept = as.numeric(df_income_change_overall),
        col = "red"
    ) +
    labs(
        x = "Total variation distance between income distributions in 2010 and 2019",
        title = "Histogram of total variation distances between income distributions across census tracts",  
        y = "Density of census tracts in class", 
        subtitle = "(Vertical red line represents the overall distribution change across Atlanta, vertical blue line represents this plus the IQR overall all"
    )

ggplot() +
    geom_sf(
        data = df_income_change_tracts %>%
            mutate(income_dist_change = pmin(income_dist_change, 0.31)), 
        aes(geometry = geometry, fill = income_dist_change),
        colour = "grey"
    ) + scale_fill_viridis(
        labels = c(0, 0.1, 0.2, "0.3+")
    ) +
    geom_sf(data = arc_counties_shp, colour = "black", fill = NA) +
    geom_sf(data = highway_arc, colour = "#000000", lty = 3, fill = NA) +
    theme(legend.position = "bottom") +
    labs(
        fill = "Total variation of income distributions",
        title = "Change in income distribution by census tract",
        subtitle = "(Data from the 5 year 2006-2010 and 2015-2019 ACS)"
    )
```

This seems pretty abstract, and so instead we'll look at some "extreme groups":

- households earning more than $200k,
- households earning less than $50k;
- households earning between $50k and $75k;

and see where the numbers of these households/proportions have been 
increasing and decreasing in number the largest. 

```{r}
df_rdincome_tracts <- df_income_tracts %>%
    filter(variable != "Total") %>%
    select(year, GEOID, variable, estimate) %>%
    mutate(variable = fct_collapse(variable, 
        "Less than $50,000" = c(
            "Less than $10,000", "$10,000 to $14,999", 
            "$15,000 to $19,999", "$20,000 to $24,999",
            "$25,000 to $29,999", "$30,000 to $34,999",
            "$35,000 to $39,999", "$40,000 to $44,999",
            "$45,000 to $49,999"
        ),
        "$50,000 to $74,999" = c("$50,000 to $59,999", "$60,000 to $74,999"),
        "$75,000 to $99,999" = c("$75,000 to $99,999"),
        "$100,000 to $149,999" = c(
            "$100,000 to $124,999", "$125,000 to $149,999"
        ),
        "$150,000 to $199,999" = c("$150,000 to $199,999"),
        "$200,000 or more" = c("$200,000 or more")
    )) %>%
    mutate(
        variable = factor(variable, ordered = TRUE, levels = income_var_labels)
    ) %>%
    group_by(year, GEOID, variable) %>%
    summarize(estimate = sum(estimate)) %>%
    left_join(df_tracts)

p2010 <- ggplot() +
    geom_sf(
        data = df_rdincome_tracts %>% filter(year == '2010'),
        aes(geometry = geometry, fill = estimate),
        colour = "grey", lwd = 0.25
    ) + scale_fill_viridis(option = "viridis", trans = 'sqrt') +
    facet_wrap(~ variable) + 
    geom_sf(data = arc_counties_shp, colour = "black", fill = NA) +
    geom_sf(data = highway_arc, colour = "#f100e5", fill = NA) +
    theme(legend.position = "bottom") + 
    labs(
        fill = "Number of households",
        title = "Number of households by income band and tract",
        subtitle = "(Data using 5-year estimates from the 2006-2010 ACS)"
    ) +
    theme(legend.key.width = unit(2, 'cm'))

p2019 <- ggplot() +
    geom_sf(
        data = df_rdincome_tracts %>% filter(year == '2019'),
        aes(geometry = geometry, fill = estimate),
        colour = "grey", lwd = 0.25
    ) + scale_fill_viridis(option = "viridis", trans = 'log10') +
    facet_wrap(~ variable) +
    geom_sf(data = arc_counties_shp, colour = "black", fill = NA) +
    geom_sf(data = highway_arc, colour = "#f100e5", fill = NA) +
    theme(legend.position = "bottom") +
    labs(
        fill = "Number of households",
        title = "Number of households by income band and tract",
        subtitle = "(Data using 5-year estimates from the 2015-2019 ACS)"
    ) +
    theme(legend.key.width = unit(2, 'cm'))

df_rdincome_change_tracts <- df_income_tracts %>%
    filter(variable != "Total") %>%
    select(year, GEOID, variable, estimate) %>%
    mutate(variable = fct_collapse(variable, 
        "Less than $50,000" = c(
            "Less than $10,000", "$10,000 to $14,999", 
            "$15,000 to $19,999", "$20,000 to $24,999",
            "$25,000 to $29,999", "$30,000 to $34,999",
            "$35,000 to $39,999", "$40,000 to $44,999",
            "$45,000 to $49,999"
        ),
        "$50,000 to $74,999" = c("$50,000 to $59,999", "$60,000 to $74,999"),
        "$75,000 to $99,999" = c("$75,000 to $99,999"),
        "$100,000 to $149,999" = c(
            "$100,000 to $124,999", "$125,000 to $149,999"
        ),
        "$150,000 to $199,999" = c("$150,000 to $199,999"),
        "$200,000 or more" = c("$200,000 or more")
    )) %>%
    mutate(
        variable = factor(variable, ordered = TRUE, levels = income_var_labels)
    ) %>%
    group_by(year, GEOID, variable) %>%
    summarize(estimate = sum(estimate)) %>%
    pivot_wider(
        names_from = "year",
        names_prefix = "estimate_",
        values_from = "estimate"
    ) %>%
    mutate(change = estimate_2019 - estimate_2010) %>%
    left_join(df_tracts)

df_rdincome_change_tracts %>%
    ggplot(aes(x = change)) + geom_histogram() +
    facet_wrap(~ variable) + 
    labs(
        x = "Change in number of households", 
        y = "Number of census tracts",
        title = "Change in number of households in different income bands across census tracts", 
        subtitle = "(Data from the 5-year ACS 2006-2010 and 2015-2019)"
    )

pminmax <- function(x, min_val, max_val) {
    return(pmin(pmax(x, min_val), max_val))
}
```

```{r}
df_rdincome_prop_tracts <- df_income_tracts %>%
    filter(variable != "Total") %>%
    select(year, GEOID, variable, estimate) %>%
    mutate(variable = fct_collapse(variable, 
        "Less than $50,000" = c(
            "Less than $10,000", "$10,000 to $14,999", 
            "$15,000 to $19,999", "$20,000 to $24,999",
            "$25,000 to $29,999", "$30,000 to $34,999",
            "$35,000 to $39,999", "$40,000 to $44,999",
            "$45,000 to $49,999"
        ),
        "$50,000 to $74,999" = c("$50,000 to $59,999", "$60,000 to $74,999"),
        "$75,000 to $99,999" = c("$75,000 to $99,999"),
        "$100,000 to $149,999" = c(
            "$100,000 to $124,999", "$125,000 to $149,999"
        ),
        "$150,000 to $199,999" = c("$150,000 to $199,999"),
        "$200,000 or more" = c("$200,000 or more")
    )) %>%
    mutate(
        variable = factor(variable, ordered = TRUE, levels = income_var_labels)
    ) %>%
    group_by(year, GEOID, variable) %>%
    summarize(estimate = sum(estimate)) %>% 
    group_by(year, GEOID) %>%
    mutate(estimate = estimate / sum(estimate)) %>%
    left_join(df_tracts)

pincome2010 <- ggplot() +
    geom_sf(
        data = df_rdincome_prop_tracts %>% filter(year == '2010'),
        aes(geometry = geometry, fill = estimate),
        colour = "grey", lwd = 0.25
    ) + scale_fill_viridis(option = "viridis", trans = 'sqrt') +
    facet_wrap(~ variable) + 
    geom_sf(data = arc_counties_shp, colour = "black", fill = NA) +
    geom_sf(data = highway_arc, colour = "#f100e5", fill = NA) +
    theme(legend.position = "bottom") + 
    labs(
        fill = "Proportion of households",
        title = "Proportion of households by income band in census tracts",
        subtitle = "(Data using 5-year estimates from the 2006-2010 ACS)"
    ) +
    theme(legend.key.width = unit(2, 'cm'))

pincome2019 <- ggplot() +
    geom_sf(
        data = df_rdincome_prop_tracts %>% filter(year == '2019'),
        aes(geometry = geometry, fill = estimate),
        colour = "grey", lwd = 0.25
    ) + scale_fill_viridis(option = "viridis", trans = 'log10') +
    facet_wrap(~ variable) +
    geom_sf(data = arc_counties_shp, colour = "black", fill = NA) +
    geom_sf(data = highway_arc, colour = "#f100e5", fill = NA) +
    theme(legend.position = "bottom") +
    labs(
        fill = "Proportion of households",
        title = "Proportion of households by income band in census tracts",
        subtitle = "(Data using 5-year estimates from the 2015-2019 ACS)"
    ) +
    theme(legend.key.width = unit(2, 'cm'))

df_rdincome_prop_change_tracts <- df_income_tracts %>%
    filter(variable != "Total") %>%
    select(year, GEOID, variable, estimate) %>%
    mutate(variable = fct_collapse(variable, 
        "Less than $50,000" = c(
            "Less than $10,000", "$10,000 to $14,999", 
            "$15,000 to $19,999", "$20,000 to $24,999",
            "$25,000 to $29,999", "$30,000 to $34,999",
            "$35,000 to $39,999", "$40,000 to $44,999",
            "$45,000 to $49,999"
        ),
        "$50,000 to $74,999" = c("$50,000 to $59,999", "$60,000 to $74,999"),
        "$75,000 to $99,999" = c("$75,000 to $99,999"),
        "$100,000 to $149,999" = c(
            "$100,000 to $124,999", "$125,000 to $149,999"
        ),
        "$150,000 to $199,999" = c("$150,000 to $199,999"),
        "$200,000 or more" = c("$200,000 or more")
    )) %>%
    mutate(
        variable = factor(variable, ordered = TRUE, levels = income_var_labels)
    ) %>%
    group_by(year, GEOID, variable) %>%
    summarize(estimate = sum(estimate)) %>% 
    group_by(year, GEOID) %>%
    mutate(estimate = estimate / sum(estimate)) %>%
    ungroup() %>%
    pivot_wider(
        names_from = "year",
        names_prefix = "estimate_",
        values_from = "estimate"
    ) %>%
    mutate(change = estimate_2019 - estimate_2010) %>%
    left_join(df_tracts)

df_rdincome_prop_change_tracts %>%
    ggplot(aes(x = change)) + geom_histogram() +
    facet_wrap(~ variable)

p1 <- ggplot() +
    geom_sf(
        data = df_rdincome_prop_change_tracts %>% 
            mutate(change = pminmax(change, -0.2, 0.2)),
        aes(geometry = geometry, fill = change),
        colour = "grey", lwd = 0.25
    ) + scale_fill_viridis(
        option = "turbo",
        labels = c("<= -0.2", -0.1, 0, 0.1, "0.2 <=")
    ) +
    facet_wrap(~ variable) + 
    geom_sf(data = arc_counties_shp, colour = "black", fill = NA) +
    geom_sf(data = highway_arc, colour = "black", lty = 3, fill = NA) +
    theme(legend.position = "bottom") +
    labs(
        fill = "Change in proportion of households",
        title = "Change in proportion of households in income band by tract",
        subtitle = "(Data using 5-year estimates from the 2006-2010 ACS and 2015-2019 ACS)"
    ) +
    theme(legend.key.width = unit(2, 'cm'))

p1
ggsave("img-files/income_changes_tract_income_prop.pdf")

p2 <- ggplot() +
    geom_sf(
        data = df_rdincome_prop_change_tracts %>%
            mutate(change = pminmax(change, -0.2, 0.2)) %>%
            filter(variable %in% c("Less than $50,000", "$200,000 or more")),
        aes(geometry = geometry, fill = change),
        colour = "grey", lwd = 0.25
    ) + 
    scale_fill_viridis(
        option = "turbo", 
        labels = c("<= -0.2", -0.1, 0, 0.1, "0.2 <=")
    ) +
    facet_wrap(~ variable) + 
    geom_sf(data = arc_counties_shp, colour = "black", fill = NA) +
    geom_sf(data = highway_arc, colour = "black", lty = 3, fill = NA) +
    theme(legend.position = "bottom") +
    labs(
        fill = "Change in proportion of households",
        title = "Change in proportion of households in income band by tract",
        subtitle = "(Data using 5-year estimates from the 2006-2010 ACS and 2015-2019 ACS)"
    ) +
    theme(legend.key.width = unit(2, 'cm'))

p2
ggsave("img-files/income_changes_tract_income_prop_highlow.pdf")
```

Let's zoom inside the perimeter to see what is happening here
a bit more carefully.

```{r}
p1 + coord_sf(xlim = c(-84.53, -84.2), ylim = c(33.6, 33.95), expand = FALSE)
ggsave("img-files/income_changes_tract_income_prop_itp.pdf")

p2 + coord_sf(xlim = c(-84.53, -84.2), ylim = c(33.6, 33.95), expand = FALSE) 
ggsave("img-files/income_changes_tract_income_prop_highlow_itp.pdf")
```

Let's also just look at the actual numbers of households, rather than the
change in proportions over time. (This should let us see if the effects we
are seeing are due to chance, or are of evidence of an actual pattern). 

```{r}
df_rdincome_total_change_tracts <- df_income_tracts %>%
    filter(variable != "Total") %>%
    select(year, GEOID, variable, estimate, moe) %>%
    mutate(variable = fct_collapse(variable, 
        "Less than $50,000" = c(
            "Less than $10,000", "$10,000 to $14,999", 
            "$15,000 to $19,999", "$20,000 to $24,999",
            "$25,000 to $29,999", "$30,000 to $34,999",
            "$35,000 to $39,999", "$40,000 to $44,999",
            "$45,000 to $49,999"
        ),
        "$50,000 to $74,999" = c("$50,000 to $59,999", "$60,000 to $74,999"),
        "$75,000 to $99,999" = c("$75,000 to $99,999"),
        "$100,000 to $149,999" = c(
            "$100,000 to $124,999", "$125,000 to $149,999"
        ),
        "$150,000 to $199,999" = c("$150,000 to $199,999"),
        "$200,000 or more" = c("$200,000 or more")
    )) %>%
    mutate(
        variable = factor(variable, ordered = TRUE, levels = income_var_labels)
    ) %>%
    group_by(year, GEOID, variable) %>%
    summarize(
        estimate = sum(estimate),
        moe = sqrt(sum(moe**2))
    ) %>%
    ungroup() %>%
    pivot_wider(
        names_from = "year",
        values_from = c("estimate", "moe")
    ) %>%
    mutate(
        change = estimate_2019 - estimate_2010,
        ratio = estimate_2019 / estimate_2010,
        log_ratio_moe = sqrt((moe_2019/estimate_2019)**2 + (moe_2010/estimate_2010)**2)
    ) %>%
    left_join(df_tracts)

p1_total <- ggplot() +
    geom_sf(
        data = df_rdincome_total_change_tracts %>%
        mutate(change = pminmax(change, -500, 500)),
        aes(geometry = geometry, fill = change),
        colour = "grey", lwd = 0.25
    ) + scale_fill_viridis(
        option = "turbo",
        labels = c("<= -500", -250, 0, 250, "500 <=")
    ) +
    facet_wrap(~ variable) + 
    geom_sf(data = arc_counties_shp, colour = "black", fill = NA) +
    geom_sf(data = highway_arc, colour = "black", lty = 3, fill = NA) +
    theme(legend.position = "bottom") +
    labs(
        fill = "Change in number of households",
        title = "Change in number of households in income band by tract",
        subtitle = "(Data using 5-year estimates from the 2006-2010 ACS and 2015-2019 ACS)"
    ) +
    theme(legend.key.width = unit(2, 'cm'))

p1_total
ggsave("img-files/income_changes_tract_income_net_change.pdf")

p2_total <- ggplot() +
    geom_sf(
        data = df_rdincome_total_change_tracts %>%
            mutate(change = pminmax(change, -500, 500)) %>%
            filter(variable %in% c("Less than $50,000", "$200,000 or more")),
        aes(geometry = geometry, fill = change),
        colour = "grey", lwd = 0.25
    ) + scale_fill_viridis(
        option = "turbo",
        labels = c("<= -500", -250, 0, 250, "500 <=")
    ) +
    facet_wrap(~ variable) + 
    geom_sf(data = arc_counties_shp, colour = "black", fill = NA) +
    geom_sf(data = highway_arc, colour = "black", lty = 3, fill = NA) +
    theme(legend.position = "bottom") +
    labs(
        fill = "Change in number of households",
        title = "Change in number of households in income band by tract",
        subtitle = "(Data using 5-year estimates from the 2006-2010 ACS and 2015-2019 ACS)"
    ) +
    theme(legend.key.width = unit(2, 'cm'))

p2_total
ggsave("img-files/income_changes_tract_income_net_change_highlow.pdf")
```

```{r}
p1_total + coord_sf(xlim = c(-84.53, -84.2), ylim = c(33.6, 33.95), expand = FALSE)
ggsave("img-files/income_changes_tract_income_net_change_itp.pdf")

p2_total + coord_sf(xlim = c(-84.53, -84.2), ylim = c(33.6, 33.95), expand = FALSE) 
ggsave("img-files/income_changes_tract_income_net_change_highlow_itp.pdf")
```

## Correlations between income changes in tracts

```{r}
df_rdincome_tc_corr <- df_income_tracts %>%
    filter(variable != "Total") %>%
    select(year, GEOID, variable, estimate) %>%
    mutate(variable = fct_collapse(variable,
        "band_1" = c(
            "Less than $10,000", "$10,000 to $14,999", 
            "$15,000 to $19,999", "$20,000 to $24,999",
            "$25,000 to $29,999", "$30,000 to $34,999",
            "$35,000 to $39,999", "$40,000 to $44,999",
            "$45,000 to $49,999", "$50,000 to $59,999",
            "$60,000 to $74,999"
        ),
        "band_2" = c("$75,000 to $99,999"),
        "band_3" = c(
            "$100,000 to $124,999", "$125,000 to $149,999"
        ),
        "band_4" = c("$150,000 to $199,999", "$200,000 or more")
    )) %>%
    mutate(
        variable = factor(
            variable, ordered = TRUE, 
            levels = c(
                "band_1", "band_2", "band_3", "band_4"
            )
        )
    ) %>%
    group_by(year, GEOID, variable) %>%
    summarize(estimate = sum(estimate)) %>%
    ungroup() %>%
    pivot_wider(
        names_from = "year",
        names_prefix = "estimate_",
        values_from = "estimate"
    ) %>%
    mutate(change = estimate_2019 - estimate_2010)

dft <- df_rdincome_tc_corr %>%
    select(GEOID, variable, change) %>%
    pivot_wider(names_from = variable, values_from = change) %>%
    left_join(
        df_tracts %>% 
            st_drop_geometry() %>%
            select(GEOID, class)
    )

ggplot(dft, aes(x = band_1 + band_2, y = band_4)) +
    geom_point()
```

# Migration patterns - income of people moving inside census tracts

```{r}
income_variables <- paste0("B07010_0", str_pad(1:66, 2, pad = "0"))

income_variables_df <- v11 %>%
    filter(name %in% income_variables) %>% 
    select(name, label) %>% 
    separate(
        label, 
        c('estimate', 'total', 'moved', 'income', 'band'), 
        sep = "!!"
    ) %>%
    drop_na() %>%
    select(name, moved, band) %>%
    filter(moved != "Moved from abroad") %>%
    mutate(
        in_out_state = fct_collapse(
            as_factor(moved),
            "Same house 1 year ago" = "Same house 1 year ago",
            "Moved in-state" = c(
                "Moved within same county",
                "Moved from different county within same state",
                "Moved from different state"
            ),
            "Moved from out-of-state" = c("Moved from different state")
        ),
        band = fct_collapse(
            as_factor(band),
            "$1 to $14,999" = c(
                "$1 to $9,999 or loss", "$10,000 to $14,999"
            ),
            "$15,000 to $34,999" = c(
                "$15,000 to $24,999", "$25,000 to $34,999"
            ),
            "$35,000 to $49,999" = "$35,000 to $49,999",
            "$50,000 to $74,999" = c(
                "$50,000 to $64,999", "$65,000 to $74,999"
            ),
            "$75,000 or more" = "$75,000 or more"
        )
    )

df_income_in <- map_dfr(
    years_acs1,
    ~ get_acs(
        geography = "county",
        state = "GA",
        variables = income_variables,
        year = .x,
        county = arc_counties,
        survey = "acs1",
        geometry = FALSE
    ),
    .id = "year") %>%
    mutate(county = gsub("^(.*?),.*", "\\1", NAME)) %>%
    select(-NAME) %>%
    relocate(county, .after = GEOID)

df_income_in %>%
    left_join(income_variables_df, by = c("variable" = "name")) %>%
    mutate(year = as.Date(paste(year, 1, 1, sep = "-"))) %>%
    drop_na() %>%
    group_by(year, in_out_state, band) %>% 
    summarize(
        estimate = sum(estimate),
        moe = sqrt(sum(moe**2))
    ) %>%
    filter(str_detect(in_out_state, "Moved")) %>%
    ggplot(aes(
        x = year, y = estimate, ymin = estimate - moe,
        ymax = estimate + moe, col = band, fill = band)
    ) + geom_line() +
    geom_point() +
    geom_ribbon(alpha = 0.15, show.legend = FALSE) +
    theme(legend.position = "bottom") +
    facet_wrap( ~ in_out_state, scales = "free") +
    scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
    labs(
        x = "Year", y = "Number of Individuals",
        col = "Individual Income",
        title = "Distribution of income of individuals who moved into Atlanta from out of state",
        subtitle = "Data from 1-year ACS from 2009 to 2019"
    ) +
    guides(fill = "none")

ggsave("img-files/income_out_of_state.pdf")

df_income_in_5yr <- map_dfr(
    years_acs5,
    ~ get_acs(
        geography = "county",
        state = "GA",
        variables = income_variables,
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = FALSE
    ),
    .id = "year") %>%
    mutate(county = gsub("^(.*?),.*", "\\1", NAME)) %>%
    select(-NAME) %>%
    relocate(county, .after = GEOID)

df_income_in_5yr %>%
    left_join(income_variables_df, by = c("variable" = "name")) %>%
    mutate(year = as.Date(paste(year, 1, 1, sep = "-"))) %>%
    drop_na() %>%
    group_by(year, in_out_state, band) %>%
    summarize(
        estimate = sum(estimate),
        moe = sqrt(sum(moe**2))
    ) %>%
    filter(in_out_state == "Moved from out-of-state") %>%
    ggplot(aes(
        x = year, y = estimate, ymin = estimate - moe,
        ymax = estimate + moe, col = band, fill = band)
    ) + geom_line() + 
    geom_point() + 
    geom_ribbon(alpha = 0.15, show.legend = FALSE) +
    theme(legend.position = "bottom") + 
    scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
    labs(
        x = "Year", y = "Number of Individuals", 
        col = "Individual Income",
        title = "Distribution of income of individuals who moved into Atlanta from out of state",
        subtitle = "Data from 5-year ACS from 2005-2009 to 2016-2020 (years not directly comparable)"
    ) +
    guides(fill = "none")
```

## Are individuals being forced out?

Before doing anything with the data, we note that there is nothing casual we
can say here. However, we can describe associations between different income groups;
here, we investigate whether there is a general association between the change
in those earning less than $75,000 a year, and those earning $200,000 or more in an area.

```{r}
df_income_change_corr <- df_rdincome_total_change_tracts %>% 
    select(GEOID, variable, ratio, log_ratio_moe) %>%
    filter(variable %in% c("Less than $50,000", "$200,000 or more")) %>%
    mutate(variable = fct_recode(
        variable, "lt_50k" = "Less than $50,000", "gt_200k" = "$200,000 or more"
    )) %>%
    pivot_wider(
        names_from = variable,
        values_from = c("ratio", "log_ratio_moe")
    ) %>%
    filter(
        ratio_gt_200k > 0, ratio_gt_200k < Inf,
        log_ratio_moe_lt_50k < 5, log_ratio_moe_gt_200k < 5
    )

dr <- deming(
    log(ratio_gt_200k) ~ log(ratio_lt_50k), data = df_income_change_corr,
    xstd = log_ratio_moe_lt_50k, ystd = log_ratio_moe_gt_200k
)
dr

dr2 <- deming(
    log(ratio_lt_50k) ~ log(ratio_gt_200k), data = df_income_change_corr,
    ystd = log_ratio_moe_lt_50k, xstd = log_ratio_moe_gt_200k
)
dr2
```

## Change in number of households earning less than $75,000 in Atlanta and the surrounding area 

In order to try and examine more carefully where people in the lower
income band are moving, we look at the change in households over the
entire Atlanta Metropoliton Statistical Area in the $75,000 or less
income bracket. We note that this roughly corresponds to
the [current living wage](https://livingwage.mit.edu/metros/12060)) estimate for a household in Atlanta, assuming a
two adult household with two children and one adult working.

```{r}

```


# How has the amount of rent changed in the Atlanta area from 2010 to 2020?

From the data and graph below, we can see that prior to the recession,
the majority of the amount paid per month in rent was less than $1,249
a month, whereas now a considerable number of units require at least
this much to pay. That said, this is not accounting directly
for changes in inflation; however, when looking at the median and lower
quartile monthly rent, these are typically around 20% and 10% higher
than what they were ten years ago, even when adjusting for inflation.
The same occurs in most of the counties for the upper quartile of
monthly rent.

(Note: the income adjustment factor between 2010 and 2020 is 1.19
according to the [Consumer Price Index](https://www.bls.gov/cpi/research-series/r-cpi-u-rs-home.htm))

```{r}
v20 <- load_variables(2020, "acs5", cache = TRUE)

rent_variables_2010 <- paste0("B25063_0", str_pad(1:24, 2, pad = "0"))

rent_var_labels_2010 <- v11 %>%
    filter(name %in% rent_variables_2010) %>%
    select(label) %>% as_vector()

rent_var_labels_2010 <- sapply(
    str_split(rent_var_labels_2010, "!!"), tail, 1
)

rent_var_list_2010 <- setNames(
    as.list(rent_variables_2010), rent_var_labels_2010
)

df_rent_distribution_2010 <- map_dfr(
    lst(2010),
    ~ get_acs(
        geography = "county",
        state = "GA",
        variables = rent_var_list_2010,
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = FALSE
    ),
    .id = "year") %>%
    mutate(
        county = gsub("^(.*?),.*", "\\1", NAME)
    ) %>% select(-NAME) %>% relocate(county, .after = GEOID)

rent_variables_2020 <- paste0("B25063_0", str_pad(1:27, 2, pad = "0"))

rent_var_labels_2020 <- v20 %>%
    filter(name %in% rent_variables_2020) %>%
    select(label) %>% as_vector()

rent_var_labels_2020 <- sapply(
    str_split(rent_var_labels_2020, "!!"), tail, 1
)

rent_var_list_2020 <- setNames(
    as.list(rent_variables_2020), rent_var_labels_2020
)

df_rent_distribution_2020 <- map_dfr(
    lst(2020),
    ~ get_acs(
        geography = "county",
        state = "GA",
        variables = rent_var_list_2020,
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = FALSE
    ),
    .id = "year") %>%
    mutate(
        county = gsub("^(.*?),.*", "\\1", NAME)
    ) %>% select(-NAME) %>% relocate(county, .after = GEOID)

df_rent_distribution <- bind_rows(
    df_rent_distribution_2010, df_rent_distribution_2020
)

collapsed_factors <- c(
    "Less than $750", "$750 to $999", "$1,000 to $1,249",
    "$1,250 to $1,499", "$1,500 to $1,999", "$2,000 or more"
)

df_rent_dist_summary <- df_rent_distribution %>% 
    mutate(variable = fct_collapse(
    variable,
    "Less than $750" = c(
        "Less than $100", "$100 to $149", "$150 to $199", "$200 to $249",
        "$250 to $299", "$300 to $349", "$350 to $399",
        "$400 to $449", "$450 to $499", "$500 to $549",
        "$550 to $599", "$600 to $649", "$650 to $699",
        "$700 to $749"
    ),
    "$750 to $999" = c(
        "$750 to $799", "$800 to $899", "$900 to $999"
    ),
    "$1,000 to $1,249" = c(
        "$1,000 to $1,249"
    ),
    "$1,250 to $1,499" = c(
        "$1,250 to $1,499"
    ),
    "$1,500 to $1,999" = c(
        "$1,500 to $1,999"
    ),
    "$2,000 or more" = c(
        "$2,000 or more", "$2,000 to $2,499",
        "$2,500 to $2,999", "$3,000 to $3,499",
        "$3,500 or more"
    )
)) %>% group_by(year, variable) %>%
    summarize(
        estimate = sum(estimate), moe = sqrt(sum(moe**2))
    ) %>%
    filter(variable %in% collapsed_factors) %>%
    mutate(
        variable = factor(variable, 
            ordered = "true", levels = collapsed_factors)
        )

df_rent_dist_summary

df_rent_dist_summary %>%
    ggplot(aes(x = variable, y = estimate, fill = year)) +
    geom_col(position = "dodge") + 
    geom_errorbar(
        aes(ymin = estimate - moe, ymax = estimate + moe),
        position = "dodge"
    ) + scale_fill_manual(values = cbPalette) +
    geom_text(
        aes(label = year, y = 10000), 
        position = position_dodge(width = 1),
        size = 5, hjust = 'left',
        col = "black"
    ) +
    labs(
        x = "Gross Rent", y = "Number of renter-occupied housing units", fill = "Year",
        title = "Distribution of monthly gross rent in the Atlanta area from 2010 to 2019",
        subtitle = "Data from the 5-year ACS from 2006-2010 to 2016-2020"
    ) +
    coord_flip() + 
    theme(legend.position = "none")

df_rent_lower_quartile <- map_dfr(
    lst(2010, 2020),
    ~ get_acs(
        geography = "county",
        state = "GA",
        variables = "B25057_001",
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = FALSE
    ),
    .id = "year") %>%
    mutate(
        county = gsub("^(.*?),.*", "\\1", NAME)
    ) %>% select(-NAME) %>% relocate(county, .after = GEOID) %>%
    select(county, year, estimate) %>%
    pivot_wider(
        names_from = year, values_from = estimate,
        names_prefix = "rent_"
    ) %>%
    mutate(
        rent_2010_in_2020_dollars = rent_2010 * 1.19,
        ratio_adjusting_for_inflation = rent_2020 / (rent_2010 * 1.19)
    )

df_rent_lower_quartile

df_rent_median_quartile <- map_dfr(
    lst(2010, 2020),
    ~ get_acs(
        geography = "county",
        state = "GA",
        variables = "B25058_001",
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = FALSE
    ),
    .id = "year") %>%
    mutate(
        county = gsub("^(.*?),.*", "\\1", NAME)
    ) %>% select(-NAME) %>% relocate(county, .after = GEOID) %>%
    select(county, year, estimate) %>%
    pivot_wider(
        names_from = year, values_from = estimate,
        names_prefix = "rent_"
    ) %>%
    mutate(
        rent_2010_in_2020_dollars = rent_2010 * 1.19,
        ratio_adjusting_for_inflation = rent_2020 / (rent_2010 * 1.19)
    )

df_rent_median_quartile

df_rent_upper_quartile <- map_dfr(
    lst(2010, 2020),
    ~ get_acs(
        geography = "county",
        state = "GA",
        variables = "B25059_001",
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = FALSE
    ),
    .id = "year") %>%
    mutate(
        county = gsub("^(.*?),.*", "\\1", NAME)
    ) %>% select(-NAME) %>% relocate(county, .after = GEOID) %>%
    select(county, year, estimate) %>%
    pivot_wider(
        names_from = year, values_from = estimate,
        names_prefix = "rent_"
    ) %>%
    mutate(
        rent_2010_in_2020_dollars = rent_2010 * 1.19,
        ratio_adjusting_for_inflation = rent_2020 / (rent_2010 * 1.19)
    )

df_rent_upper_quartile
```

## Correlation with change in number of households in income brackets

Hmm, suprisingly there doesn't appear to be any correlation between the change 
in the number of households in the lower income bracket as compared to the change
in rent in a particular census tract. To be honest, I find this pretty surprising...

```{r}
df_rent_lower_quartile_tracts <- map_dfr(
    lst(2010, 2019),
    ~ get_acs(
        geography = "tract",
        state = "GA",
        variables = "B25057_001",
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = FALSE
    ),
    .id = "year") %>% mutate(
        tract = gsub("^(.*?),.*", "\\1", NAME),
        county = gsub("^(.*?), (.*?),.*", "\\2", NAME)
    ) %>% 
    select(-NAME) %>%
    relocate(tract, county, .after = GEOID) %>%
    filter(tract != "Census Tract 9800") %>%
    select(GEOID, year, estimate) %>%
    pivot_wider(
        names_from = year, values_from = estimate,
        names_prefix = "rent_"
    ) %>%
    mutate(
        rent_2010_in_2019_dollars = rent_2010 * 1.175,
        ratio_adjusting_for_inflation = rent_2019 / (rent_2010 * 1.175)
    )

df_ratio_change_tracts <- df_rdincome_total_change_tracts %>%
    select(-geometry) %>%
    filter(variable %in% c("Less than $50,000")) %>%
    select(GEOID, variable, estimate_2010, estimate_2019, class) %>%
    group_by(GEOID) %>%
    mutate(
        estimate_2010 = sum(estimate_2010),
        estimate_2019 = sum(estimate_2019)
    ) %>% ungroup() %>% unique() %>%
    mutate(
        pop_ratio = estimate_2019 / estimate_2010
    )

df_tract_changes <- left_join(
    df_rent_upper_quartile_tracts, df_ratio_change_tracts
)

df_tract_changes %>% 
    ggplot(aes(x = ratio_adjusting_for_inflation, y = pop_ratio, col = class)) +
    scale_x_log10() + scale_y_log10() +
    geom_point()

cor.test(
    df_tract_changes$ratio_adjusting_for_inflation,
    df_tract_changes$pop_ratio,
    use = "complete.obs"
)
```

However, if we look at
- the ratio of households earning less than $50,000 a year between 2019 and 2010, and
- the number of renters paying less than $1,000 a month in rent between 2019 and 2010
across the tracts, we see that there is a statistically significant relationship
between the two ratios.

In performing this calculation, we removed census tracts for which the
estimated number of households paying less than $1,000 in rent was zero in
either of the time periods. We also removed ratios whose standard errors were too
high. To do so, we note that it is usually assumed that, across different periods
of time, variables are independent, and are approximately Normally distributed
with mean given by the estimate, and standard error given by the [margin of error
divided by 1.645](https://www.census.gov/programs-surveys/acs/guidance/training-presentations/acs-moe.html). In general, if $X \sim N(\mu_X, \sigma_X^2)$
and $Y \sim N(\mu_Y, \sigma_Y^2)$ are independent, then we can approximate
the standard error of $\log(X/Y)$ by

$$\sqrt( \frac{\sigma_X^2}{\mu_X^2} + \frac{\sigma_Y^2}{\mu_Y^2}).$$

For each census tract, we calculate the above quantity, and remove any
census tract for which the standard error of the above quantity was greater than
five. After these two steps, we remained with 500 of the 783 original census tracts.
With these variables, we then perform a [Deming regression](https://en.wikipedia.org/wiki/Deming_regression) of the two ratios against each other, which accounts for the variability
in both of the observations. 

```{r}
df_rent_distribution_2010_tract <- map_dfr(
    lst(2010),
    ~ get_acs(
        geography = "tract",
        state = "GA",
        variables = rent_var_list_2010,
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = FALSE
    ),
    .id = "year") %>% mutate(
        tract = gsub("^(.*?),.*", "\\1", NAME),
        county = gsub("^(.*?), (.*?),.*", "\\2", NAME)
    ) %>% 
    select(-NAME) %>%
    relocate(tract, county, .after = GEOID) 

df_rent_distribution_2019_tract <- map_dfr(
    lst(2019),
    ~ get_acs(
        geography = "tract",
        state = "GA",
        variables = rent_var_list_2020,
        year = .x,
        county = arc_counties,
        survey = "acs5",
        geometry = FALSE
    ),
    .id = "year") %>% mutate(
        tract = gsub("^(.*?),.*", "\\1", NAME),
        county = gsub("^(.*?), (.*?),.*", "\\2", NAME)
    ) %>% 
    select(-NAME) %>%
    relocate(tract, county, .after = GEOID) 

df_rent_distribution_tract <- bind_rows(
    df_rent_distribution_2010_tract, df_rent_distribution_2019_tract
)

df_rent_dist_summary_tract <- df_rent_distribution_tract %>% 
    mutate(variable = fct_collapse(
    variable,
    "Less than $750" = c(
        "Less than $100", "$100 to $149", "$150 to $199", "$200 to $249",
        "$250 to $299", "$300 to $349", "$350 to $399",
        "$400 to $449", "$450 to $499", "$500 to $549",
        "$550 to $599", "$600 to $649", "$650 to $699",
        "$700 to $749"
    ),
    "$750 to $999" = c(
        "$750 to $799", "$800 to $899", "$900 to $999"
    ),
    "$1,000 to $1,249" = c(
        "$1,000 to $1,249"
    ),
    "$1,250 to $1,499" = c(
        "$1,250 to $1,499"
    ),
    "$1,500 to $1,999" = c(
        "$1,500 to $1,999"
    ),
    "$2,000 or more" = c(
        "$2,000 or more", "$2,000 to $2,499",
        "$2,500 to $2,999", "$3,000 to $3,499",
        "$3,500 or more"
    )
)) %>% group_by(year, variable, GEOID) %>%
    summarize(
        estimate = sum(estimate), moe = sqrt(sum(moe**2))
    ) %>%
    filter(variable %in% collapsed_factors) %>%
    mutate(
        variable = factor(variable, 
            ordered = "true", levels = collapsed_factors)
        )

df_rent_change_dist_summary <- df_rent_dist_summary_tract %>%
    filter(variable %in% c("Less than $750", "$750 to $999")) %>%
    select(year, GEOID, variable, estimate, moe) %>%
    group_by(year, GEOID) %>%
    summarize(
        estimate = sum(estimate),
        moe = sqrt(sum(moe**2))
    ) %>%
    ungroup() %>%
    pivot_wider(
        names_from = year, values_from = c(estimate, moe),
        names_prefix = "lt_750_"
    ) %>%
    mutate(
        ratio = estimate_lt_750_2019 / estimate_lt_750_2010,
        log_ratio_moe = sqrt((moe_lt_750_2019/estimate_lt_750_2019)**2 + (moe_lt_750_2010/estimate_lt_750_2010)**2)
    ) %>% left_join(
        df_rdincome_total_change_tracts %>% 
            select(-geometry) %>%
            filter(variable == 'Less than $50,000'), by = "GEOID"
    )

df_rent_change_dist_summary %>%
    filter(ratio.x > 0, ratio.x < Inf) %>%
    filter(log_ratio_moe.x < 5, log_ratio_moe.y < 5) %>%
    ggplot(aes(x = ratio.x, y = ratio.y)) + geom_point() +
    scale_x_log10() + scale_y_log10() + 
    labs(
        x = "Ratio of occupied rental units with gross rent less than $1,000 a month in rent from 2019 to 2010",
        y = "Ratio of households earning less than $50,000 from 2019 to 2010"
    )

df_rent_cor_test <- df_rent_change_dist_summary %>%
    filter(ratio.x > 0, ratio.x < Inf) %>%
    filter(log_ratio_moe.x < 5, log_ratio_moe.y < 5) 

# Not accounting for varition between multiple variables
cor.test(
    log(df_rent_cor_test$ratio.x), log(df_rent_cor_test$ratio.y),
    use = "complete.obs"
)

require('deming')
dr <- deming(
    log(ratio.y) ~ log(ratio.x), data = df_rent_cor_test,
    xstd = log_ratio_moe.x, ystd = log_ratio_moe.y
)
dr
```

# Incorporating land-use change data

- Extract the land change data from the land use dataset for two of the
years 

- Compute correlation with land use + 
    - gain in households with income in the high brackets 
    - gain in households with income in the low brackets
    - income change in general

- From the satellite imagery, look at 
areas of interest in 2010 as compared to as recently as possible,
see what changes are there
    - The place in Forsyth County
    - The place "in the wedge"
    - The place at the intersection of DeKalb and Fulton
    - 11 o-clock in the perimeter




